Re-run, Repeat, Reproduce, Reuse, Replicate:\\Transforming Code into Scientific Contributions

### 2017-08-09 draft.tex
rearranged reproducible, added git tracking, other minor edits

Executed, this program would display:
\begin{code}{Output}
-1  0 -1 -2 -1 -2 -3 -2 -1 -2 # with the steps being -1, +1, -1, -1, +1, -1, -1, +1, +1, -1
\end{code}

@@

Because any dependency of your program---to the most basic one, the language itself---can change its behavior from one version to another, executability (R$^1$) and determinism (R$^2$) are necessary but not sufficient for reproducibility. The exact execution environment used to produce the results must also be specified---rather than the broadest set of environments where the code can be effectively run. In other words, assertions such as "the results were obtained with CPython 3.6.1" are more valuable, in a scientific context, than "the program should work with Python 3.x and above".\\

%FIXME: this is not a good-enough paragraph
With the increasing complexity of our computational stacks, retrieving and deciding what is pertinent (CPU architecture? Operating system version? Endianness?) can be difficult, however. A best effort must be made.\\

Having precisely described the execution environment, reproducibility is \emph{technically} obtained. However, \emph{reproducibility in practice} requires more than that.\\

The R$^2$ program uses a random seed but does not keep a trace of it except in the code. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used, and would need to iterate through all possible random seeds, an impossible task in practice. More generally, results files should always come alongside their context, i.e. an exhaustive list of the parameters used and a description of the execution environment, as the R$^3$ code does.\\

% scripts to generates figures
The code should also contain all the scripts that were used to produce the figures. While it may be considered extraneous, code for plotting figures may contain non-trivial statistics code, and is actually just as vulnerable to bugs. With a code that is re-runnable, repeatable and contains such scripts, a user can quickly check that published figures can be reproduced.\\

% data behind graphs
However, figures by themselves are imprecise and cumbersome, and sometimes plain unusable, when it comes to checking that the reproduced data match the one published in the article. A published figure should always be accompanied by its underlying data to allow straightforward numeric comparisons.\\

@@

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 4:}} Re-runnable, repeatable, reproducible random walk (R$^3$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C. Y. Benureau
# Release under the BSD 2-clause license
# Tested with CPython 3.6.1 / macOS 10.12.4 / 64 bits architecture
import platform, datetime
import random

def provenance(): # Get info about the execution environment
    return {'python'   : {'implementation': platform.python_implementation(),
                          'version'       : platform.python_version_tuple(),
                          'compiler'      : platform.python_compiler(),
                          'branch'        : platform.python_branch(),
                          'revision'      : platform.python_revision()},
            'platform' : platform.platform(),
            'timestamp': datetime.datetime.utcnow().isoformat()+'Z'}  # Z stands for UTC

def walk():
ea    promenade, total = [], 0
    for i in range(10):
        step = +1 if random.uniform(-1,+1) > 0 else -1
        total += step
        promenade.append(total)
    return promenade

@@

% Summary
To recap, reproducibility implies re-runnability and repeatability and availability, yet imposes additional conditions. Dependencies and platforms must be described as precisely and a specifically as possible. Parameters values and inputs should be an integral part of the result files. The data and scripts behind the graphs must be published. Unit tests are a good way to embed self-diagnostics of reproducibility in the code. Reproducibility is hard, yet tremendously necessary.

@@

For you, first. Because the you now and the you in two years are two different persons. Details on how to use the code, its limitations, its quirks, may be present to your mind now, but will probably escape you in six months \parencite{Donoho:2009}. Here, comments and documentation can make a significant difference. In fact, a source code reflects the results of all the decisions that were made to create it, but not the reasons behind the decisions. In science, where the method and its justification matter as much as the results, those reasons are precious knowledge. In that context, a comment on how a given parameter was chosen (optimization, experimental data, educated guess), why a library was chosen over another (conceptual or technical reasons?) is valuable information.\\

Reusability of course directly benefits other researchers from your team and outside of it. The easier it is to use your code, the lower the threshold is for other to study, modify and extend it. Scientists constantly face the constraint of time: if a model is available, documented, and can be installed, run and understood all in a few hours, it will be preferred over another that would require weeks to reach the same stage. Moreover, scientists prefer to base their work on results they can verify. A reproducible and reusable code offers a platform both validated and easy-to-use, fostering the development of derivative works by other researchers. Those derivative works contribute to the impact of your original contribution.\\

@@

Despite all this, reusability is often overlooked, and it is not hard to see why. Scientists are rarely trained in software engineering, and reusability can represent an expensive endeavour if undertaken as an afterthought, for little tangible short-term benefits, for a codebase that might, after all, see only a single use. And, in fact, reusability is not as indispensable a requirement as the other ones presented here.\\

Yet, some simple measures can tremendously increase reusability, and at the same time strengthen reproducibility and re-runnability.\\

@@

Similarly, code behavior should be changed by commenting/uncommenting code \citep{Wilson:2017}. Modification of the behavior of the code, required when different experiments examine slightly different conditions, should always be explicitly set through parameters accessible to the end-user. This improves reproducibility in two ways: it allows those conditions to be explicitly tracked in the result files since they contain the parameters used. It also allows to define separate scripts to run or configuration files to load to produce each of the figures of the published paper. With a documention explaining which script or configuration file correspond to which experiment, reproducing the different figures becomes straightforward.\\
%FIXME repeating what was said in reproducible.

@@

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 5:}} Re-runnable, repeatable, reproducible, reusable random walk (R$^4$)}
  # Copyright (c) 2017 Nicolas P. Rougier and Fabien C. Y. Benureau
  # Release under the BSD 2-clause license
  # Tested with CPython 3.6.1 / macOS 10.12.4 / 64 bits architecture
  import platform, datetime, json
  import random

  def provenance():
      """Retrieve data about the execution environment."""
      return {'python'   : {'implementation': platform.python_implementation(),
                            'version'       : platform.python_version_tuple(),
                            'compiler'      : platform.python_compiler(),
                            'branch'        : platform.python_branch(),
                            'revision'      : platform.python_revision()},
              'platform' : platform.platform(),
              'timestamp': datetime.datetime.utcnow().isoformat()+'Z'}  # Z stands for UTC

  def walk(n, seed):
      """Return a `n` steps random walk.

      `seed` is used to initialized the Mersene Twister random number generator."""
      random.seed(seed)
      rndwalk, total = [], 0
      for i in range(10):
          step = +1 if random.uniform(-1,+1) > 0 else -1
          total += step
          rndwalk.append(total)
      return rndwalk

  # Unit test
  assert walk(10, 1) == [-1, 0, 1, 0, -1, -2, -1, 0, -1, -2]

  def generate_results(n, seed, filename='results-R4'):
      """Generate results and save them to disk as a JSON file"""
      w = walk(n, seed)

      results = {'data': w, 'n': n, 'seed': seed, 'provenance': provenance()}
      with open(filename + '.json', 'w') as fd:
          json.dump(results, fd)
      return results

  if __name__ == '__main__':
      print(generate_results(10, 1)['data'])
\end{code}
\end{minipage}\\

@@

% errors may always remain: reproducibility does not make reproducibility useless.
But even when those details are present, the best efforts may fall short because
an oversight, a typo or a difference between what is evident for the writer
and for the reader \citep{Mesnard:2016}. In that case, a reproducible code is a crucial backup. Hence,
replicability does not negate the necessity of reproducibility. To illustrate
this, let us consider what could be the textual description of the random
walker (as it would be written in an article describing it):
%
\begin{quotation}
{\em The model uses the Mersene Twister generator initialized with the seed 0. At each iteration, a uniform number between -1 (included) and +1 (excluded) is drawn and the sign of the result is used for generating a positive or negative step.}
\end{quotation}
%
This description, while somewhat precise, forgoes---as it is common---the initialization of the variables (here the starting value of the walk ({\tt 0})), and the technical details about which implementation of the RNG is used.\\

@@

%FIXME: I feel like this paragraph is openning me to attack in a way that is probably unecessary.
Striving, as we do here, for a perfect quantitative match may seem unnecessary. Yet, in replication projects, quantitative comparisons are a powerful way to verify that the behavior has been reproduced: they eliminate such a quantity of possible errors that they can act, in practice, as \emph{good enough} test of replication. They are also particularly helpful to track exactly where the code of a tentative replication fails to reproduce the results that were published.\\

@@

But there are still more subtle problems with the description given above.
If we look more closely at it, we can realize
that nothing is said about the specific case of {\tt 0} when generating a step.
Do we have to consider {\tt 0} to be a positive or a negative step? Without
further information and without the original code, it is up to the reader to
decide. Likewise, the description is ambiguous regarding the first element of the walk. Is the initialization value included (it was not in our codes so far)? This slight difference changes the statistics, especially of short runs.\\

There is no better way to ferret out the ambiguities, big and small, of an
article than to replicate it. This is one of the reasons why the ReScience
journal \cite{Rougier:2017} has been created. This journal {\em targets
  computational research and encourages the explicit replication of already
  published research, promoting new and open-source implementations in order to
  ensure that the original research is reproducible}.\\

@@

%FIXME: can still be improved
Through a small random walk example using Python, we've seen a variety of
pitfalls that are easy to remain unaware of while writing scientific code. The
R$^0$ form, while easy to write and simple to understand is not a good
scientific code. Adding a single comment, we transformed this code into the
R$^1$ version (rerunnable) that partially solved the common dependencies
problem and will allow us to re-run the program two months or two years from
now. In the meantime, the R$^2$ version (repeatable) guarantees us (to some
extent) we can faithfully repeat our experiment from one run to the other.  Of
course, the full-fledge R$^4$ (reusable) form is much preferred but it requires
a fair amount of work that might be too much of a burden. R$^3$ (reproducible)
form seems to be an acceptable compromise and should be accepted as the minimum
scientific standard \citep{Wilson:2017}. This means this should be actually
checked by reviewers and publishers when code is part of a work worth being
published. But it's hardly the case today.\\

@@

Compared to psychology or biology, the replication issues of computational works have reasonable and efficient solutions. But making sure that these solutions are adopted will not be solved by articles such as this one. Just like in psychology and biology, we have to modify the incentive of the researchers, by adopting exigences, enforced domain-wide, of what constitutes an acceptable scientific computational work.

>>>>

Executed, this program would display:
\begin{code}{Output}
-1 0 -1 -2 -1 -2 -1 -2 -3 -4 # with the steps being -1, +1, -1, -1, +1, -1, +1, -1, -1,  -1
\end{code}

@@

% precise execution environment recording
Because any dependency of a program---to the most basic one, the language itself---can change its behavior from one version to another, executability (R$^1$) and determinism (R$^2$) are necessary but not sufficient for reproducibility. The exact execution environment used to produce the results must also be specified---rather than the broadest set of environments where the code can be effectively run. In other words, assertions such as "the results were obtained with CPython 3.6.1" are more valuable, in a scientific context, than "the program should work with Python 3.x and above". With the increasing complexity of computational stacks, retrieving and deciding what is pertinent (CPU architecture? operating system version? endianness?) might be non-trivial. A good rule of thumb is to include more information than necessary rather than not enough.\\

% if code changes, parameters are lost
Recording the execution environment is not enough though. The R$^2$ program uses a random seed but does not keep a trace of it except in the code. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used, and would need to iterate through all possible random seeds, an impossible task in practice\footnote{Here, with $2^{10}$ possibilities for a 10-step random walk, a seed matching the results could obviouly be found. For instance, seed 1123581321 matches the results of R$^0$ with Python 2.7. This is not possible for a 100-step walk.}.

% 1. parameters and 2. code version with results
From this observation, we conclude two things. First, results files should always come alongside their context, i.e. an exhaustive list of the parameters used as well as a precise description of the execution environment, as the R$^3$ code does. Second, the version of the code itself should be recorded as part of the provenance data. It is common for results to have been generated by different versions of the code. Ideally, all results should originate from the same (and last) version of the code. But for long or expensive computations, this may not be feasible. In that case, the result files should contains the version of the code that was used to produce it. This information can be obtained from the version control software. This also allow, if some errors are found and corrected after some results have been obtained, to identify which ones should be recomputed. In R$^3$, the code records the current git hash, and wether the repository holds uncommited changes when the computation starts.\\

% scripts to generates figures & data behind graphs
The code should also contain all the scripts that were used to produce the figures. Plotting code may seem mundane, but it is as vulnerable as any other piece of the code to bugs and errors. With a code that is re-runnable, repeatable and contains such scripts, a user can quickly check that published figures can be reproduced. However, figures by themselves are imprecise and cumbersome, and sometimes plain unusable, when it comes to checking that the reproduced data match the one published in the article. A published figure should always be accompanied by its underlying data to allow straightforward numeric comparisons.\\

@@

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 4:}} Re-runnable, repeatable, reproducible random walk (R$^3$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C. Y. Benureau
# Release under the BSD 2-clause license
# Tested with CPython 3.6.1 / macOS 10.12.4 / 64 bits architecture
import platform, datetime
import random
import git

def git_info(): # Retrieve the code version
    repo = git.Repo(os.path.dirname(__file__), search_parent_directories=True)
    return {'hash': repo.head.object.hexsha,
            'dirty': repo.is_dirty()}

def provenance(): # Get info about the execution environment
    return {'python'   : {'implementation': platform.python_implementation(),
                          'version'       : platform.python_version_tuple(),
                          'compiler'      : platform.python_compiler(),
                          'branch'        : platform.python_branch(),
                          'revision'      : platform.python_revision()},
            'platform' : platform.platform(),
            'git_info' : git_info(),
            'timestamp': datetime.datetime.utcnow().isoformat()+'Z'}  # Z stands for UTC

def walk():
    promenade, total = [], 0
    for i in range(10):
        step = +1 if random.uniform(-1,+1) > 0 else -1
        total += step
        promenade.append(total)
    return promenade

@@

% Summary
To recap, reproducibility implies re-runnability and repeatability and availability, yet imposes additional conditions. Dependencies and platforms must be described as precisely and a specifically as possible. Parameters values and inputs should accompany the result files. The data and scripts behind the graphs must be published. Unit tests are a good way to embed self-diagnostics of reproducibility in the code. Reproducibility is hard, yet tremendously necessary.

@@

For you, first. Because the you now and the you in two years are two different persons. Details on how to use the code, its limitations, its quirks, may be present to your mind now, but will probably escape you in six months \parencite{Donoho:2009}. Here, comments and documentation can make a significant difference. In fact, a source code reflects the results of all the decisions that were made to create it, but not the reasons behind those decisions. In science, where the method and its justification matter as much as the results, those reasons are precious knowledge. In that context, a comment on how a given parameter was chosen (optimization, experimental data, educated guess), why a library was chosen over another (conceptual or technical reasons?) is valuable information.\\

Reusability of course directly benefits other researchers from your team and outside of it. The easier it is to use your code, the lower the threshold is for other to study, modify and extend it. Scientists constantly face the constraint of time: if a model is available, documented, and can be installed, run and understood all in a few hours, it will be preferred over another that would require weeks to reach the same stage. A reproducible and reusable code offers a platform both \emph{verifiable} and easy-to-use, fostering the development of derivative works by other researchers on solid fundations. Those derivative works contribute to the impact of your original contribution.\\

@@

Despite all this, reusability is often overlooked, and it is not hard to see why. Scientists are rarely trained in software engineering, and reusability can represent an expensive endeavour if undertaken as an afterthought, for little tangible short-term benefits, for a codebase that might, after all, see only a single use. And, in fact, reusability is not as indispensable a requirement as the other ones presented here. Yet, some simple measures can tremendously increase reusability, and at the same time strengthen reproducibility and re-runnability.\\

@@

Similarly, code behavior should not be changed by commenting/uncommenting code \citep{Wilson:2017}. Modification of the behavior of the code, required when different experiments examine slightly different conditions, should always be explicitly set through parameters accessible to the end-user. This improves reproducibility in two ways: it allows those conditions to be explicitly tracked in the result files since they contain the parameters used, and it allows to define separate scripts to run or configuration files to load to produce each of the figures of the published paper. With a documention explaining which script or configuration file correspond to which experiment, reproducing the different figures becomes straightforward.\\

@@

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 5:}} Re-runnable, repeatable, reproducible, reusable random walk (R$^4$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C. Y. Benureau
# Release under the BSD 2-clause license
# Tested with CPython 3.6.1 / macOS 10.12.4 / 64 bits architecture
import platform, datetime, json
import random
import git

def git_info():
    """Retrieve the current git hash, and if the repository has uncommited changes."""
    repo = git.Repo(os.path.dirname(__file__), search_parent_directories=True)
    return {'hash': repo.head.object.hexsha,
            'dirty': repo.is_dirty()}

def provenance():
    """Get info about the execution environment and the code."""
    return {'python'   : {'implementation': platform.python_implementation(),
                          'version'       : platform.python_version_tuple(),
                          'compiler'      : platform.python_compiler(),
                          'branch'        : platform.python_branch(),
                          'revision'      : platform.python_revision()},
            'platform' : platform.platform(),
            'git_info' : git_info(),
            'timestamp': datetime.datetime.utcnow().isoformat()+'Z'}  # Z stands for UTC

def walk(n, seed):
    """Return a `n` steps random walk.
    `seed` is used to initialize the Mersene Twister random number generator."""
    random.seed(seed)
    rndwalk, total = [], 0
    for i in range(10):
        step = +1 if random.uniform(-1,+1) > 0 else -1
        total += step
        rndwalk.append(total)
    return rndwalk

def generate_results(n, seed, filename='results-R4'):
    """Generate results and save them to disk as a JSON file"""
    w = walk(n, seed)
    results = {'data': w, 'n': n, 'seed': seed, 'provenance': provenance()}
    with open(filename + '.json', 'w') as fd:
        json.dump(results, fd)
    return results

if __name__ == '__main__':
    # Unit test checking reproducibility
    assert walk(10, 1) == [-1, 0, 1, 0, -1, -2, -1, 0, -1, -2]

    print(generate_results(10, 1)['data'])
\end{code}
\end{minipage}\\

@@

% errors may always remain: reproducibility does not make reproducibility useless.
But even when those details are present, the best efforts may fall short because
an oversight, a typo or a difference between what is evident for the writer
and for the reader \citep{Mesnard:2016}. In that case, a reproducible code is a crucial backup. Hence,
replicability does not negate the necessity of reproducibility. To illustrate
this, let us consider what could be the textual description of the random
walker (as it would be written in an article describing it):
%
\begin{quotation}
{\em The model uses the Mersene Twister generator initialized with the seed 0. At each iteration, a uniform number between -1 (included) and +1 (excluded) is drawn and the sign of the result is used for generating a positive or negative step.}
\end{quotation}
%
This description, while somewhat precise, forgoes---as it is common---the initialization of the variables (here the starting value of the walk: {\tt 0}), and the technical details about which implementation of the RNG is used.\\

@@

%FIXME: I feel like this paragraph is openning me to attack in a way that is probably unecessary.
Striving, as we do here, for a perfect quantitative match may seem unnecessary. Yet, in replication projects, quantitative comparisons are an effective way to verify that the behavior has been reproduced. Moreover, they are particularly helpful to track exactly where the code of a tentative replication fails to reproduce the published results.\\

@@

But there are still more subtle problems with the description given above.
If we look more closely at it, we can realize
that nothing is said about the specific case of {\tt 0} when generating a step.
Do we have to consider {\tt 0} to be a positive or a negative step? Without
further information and without the original code, it is up to the reader to
decide. Likewise, the description is ambiguous regarding the first element of the walk. Is the initialization value included (it was not in our codes so far)? This slight difference affects statistics, especially for short runs.\\

There is no better way to ferret out the ambiguities, big and small, of an
article than to replicate it. This is one of the reasons why the ReScience
journal \citep{Rougier:2017} has been created. This journal {\em targets
  computational research and encourages the explicit replication of already
  published research, promoting new and open-source implementations in order to
  ensure that the original research is reproducible}.\\

@@

%FIXME: can still be improved
Through a small random walk example implemented in Python, we've seen that writing good scientific code is not trivial. The
R$^0$ form, while easy to write and simple to understand is not a good
scientific code. Adding a single comment, we transformed this code into the
R$^1$ version (rerunnable) that partially solved the common dependencies
problem and will allow us to re-run the program two months or two years from
now. In the meantime, the R$^2$ version (repeatable) guarantees us (to some
extent) we can faithfully repeat our experiment from one run to the other.  Of
course, the full-fledge R$^4$ (reusable) form is much preferred but it requires
a fair amount of work that might be too much of a burden. R$^3$ (reproducible)
form seems to be an acceptable compromise and should be accepted as the minimum
scientific standard \citep{Wilson:2017}. This means this should be actually
checked by reviewers and publishers when code is part of a work worth being
published. But it's hardly the case today.\\

@@

Compared to psychology or biology, the replication issues of computational works have reasonable and efficient solutions. But making sure that these solutions are adopted will not be solved by articles such as this one. Just like in psychology and biology, we have to modify the incentive of the researchers, by adopting exigences, enforced domain-wide, on what constitutes an acceptable scientific computational work.
